\section{DNA Microarrays}

The first human genome sequence, completed in 2001, ushered in a new era of
biology.  This effort accelerated the development of biochemical assay and
information technologies.  As a result, biologists are now able to ask
questions that were previously considered intractable.  One example of a
breakthrough in assay technology resulting from genome sequencing efforts is
the DNA microarray, a high-throughput measurement device which enables
individual scientists to rapidly and simultaneously interrogate the RNA
concentration levels of virtually all genes in the human genome for a single
biological source.  The microarray is one of the most commonly used of
post-genomic high-throughput assay technologies, as is indicated by the number
of abstracts in PubMed containing the word ``microarray''.  It has been widely
adopted in the biological sciences and been used to address a large number of
research questions across a diverse set of subdomains in biological research.

Experiments that collect data using microarray technology can be divided into
several distinct categories, based on their overall goal and tactics employed
in analysis of the data.  These include: classification problems (supervised
learning) \cite{Dudoit2003ICM}, clustering problems (unsupervised learning)
\cite{Azuaje2003cge,Stanford2003cac}, gene co-regulation and function
identification studies \cite{PMID_12413821}, differential gene expression
studies \cite{PMID_15843092,PMID_15867208}, time-course and dose-response
studies \cite{PMID_12443997}, gene pathway and regulatory network studies
\cite{PMID_16216773,PMID_16825123}, drug discovery and toxicology studies
\cite{PMID_16512775,PMID_16700885,PMID_16880944,PMID_17195470}, clinical
diagnostics \cite{PMID_16918486}, and sequence-variation studies
\cite{PMID_17265721}.

A successful microarray experiment is one which is able to provide insight into
the hypothesis for which it was designed.  Insights gained by experimentation
are hard-earned, and require careful planning and execution of multiple key
steps.  The initial steps of feature selection, microarray design, microarray
fabrication, experiment design, and biological sample processing, and image
acquisition are described generally in Sections
\ref{ArrayDesign}-\ref{Hybridization}.  A more exhaustive discussion of
protocol surrounding these steps is given in \cite{wit2004}.  Following their
acquisition, images are quantified, stored, distributed, analyzed, and
analytical results shared.  Challenges and findings related to these latter
procedures are the subject of \emph{\dbthesis}.  They are thoroughly discussed
in Sections \ref{Data Processing}-\ref{Informatics}.  The entire process, from
microarray design to data analysis has been formally described as an object
model known as the Microarray Gene Expression Object Model (MAGE-OM)
\cite{mage}.  In this document I have borrowed vocabulary defined in the
MAGE-OM for clarity and consistency.  The MAGE-OM itself is discussed in
Section \ref{MAGE}.

\section{DNA Microarray Fabrication and Assay Protocols}
\label{Assay}
\input{section_dna_microarray_assay}

\section{DNA Microarray Experiment Design}
\label{Experiment Design}
\input{section_dna_microarray_experiments}

\section{DNA Microarray Data Processing}
\label{Data Processing}
\input{section_dna_microarray_data_processing}

\section{DNA Microarray Information Systems}
\label{Informatics}
\input{section_dna_microarray_informatics}



\subsection{Northern Blot}
\subsection{Oligonucleotide Synthesis}
\subsection{Robotics}
\subsection{Microfluidics}
\subsection{Photolithography}
\section{Microarray Assay Overview}
A successful microarray experiment is one which is able to provide insight into
the hypothesis for which it was designed.  Insights gained by experimentation
are hard-earned, and require careful planning and execution of multiple key
steps.

The initial steps of feature selection, microarray design, microarray
fabrication, experiment design, and biological sample processing, and image
acquisition are described generally in Sections
\ref{ArrayDesign}-\ref{Hybridization}.  A more exhaustive discussion of
protocol surrounding these steps is given in \cite{wit2004}.  Following their
acquisition, images are quantified, stored, distributed, analyzed, and
analytical results shared.  Challenges and findings related to these latter
procedures are the subject of \emph{\dbthesis}.  They are thoroughly discussed
in Sections \ref{Data Processing}-\ref{Informatics}.

The entire process, from microarray design to data analysis has been formally
described as an object model known as the Microarray Gene Expression Object
Model (MAGE-OM) \cite{mage}.  In this document I have borrowed vocabulary
defined in the MAGE-OM for clarity and consistency.  The MAGE-OM itself is
discussed in Section \ref{MAGE}.

\subsection{Microarray Design}
\label{ArrayDesign}

The first steps in performing a microarray experiment is deciding which genes
to assay on the microarray, how many genes to add, and how to arrange the genes
on the microarray.  These steps are collectively known as \emph{microarray
design}.

Historically, limitations in manufacturing technology were very restrictive,
and limited the number of genes that could be assayed simultaneously.  Current
technology is sufficiently advanced to allow all genes to be simultaneously
assayed for most organisms, so the question of which and how many genes to add
is no longer relevant.  However, as anyone who has performed and experiment
will tell you, the results obtained from the same experimental procedure
performed twice will yield different results.  This also holds true for
measurements made in microarray measurements.  There are many potential sources
of variance in the gene measurements, and a good array design measures as much
of this measurement variance as possible.

Estimation of variance on a microarray essentially means making multiple
measurements of the same gene.  Each gene can be measured at multiple distinct
locations along its sequence.  Further, each location can be measured multiple
times, Further location replicates can be uniformly distributed across the
array surface to compensate for local perturbations on the array surface
itself.

\subsection{Fabrication}
\label{Fabrication}

Once a microarray has been designed (Section \ref{ArrayDesign}) it can be
manufactured.  Fabrication and microarray design are interdependent, in that
improvements in manufacturing techniques allow for the fabrication of more
powerful devices.  Thus, this area of microarray technology is evolving
especially rapidly and any specific details of the synthesis we can describe
here will be outdated before this document is published.  For this reason, I
will only discuss the process of fabrication in general.  A historical
progression of technological advances in fabrication is given in Section
\ref{History}.

\section{Infrastructure}
\label{Infrastructure}

The study of biology is rapidly evolving.  Bottom-up approaches, in which data
are produced individually and slowly are being replaced by high-throughput
methods that produce, in parallel, large volumes of high-dimensional,
high-resolution measurements.  Recent and emerging technologies such as DNA
microarrays, mass spectrometry, and next-generation DNA sequencing enable the
biologist to study the molecular details of entire organisms rather than just a
few genes.  This transition has created new opportunities for discovery in
collaboration with engineers, mathematicians, statisticians, and computer
scientists.  At the same time, this transition necessitates an overhaul of the
computational infrastructure used by biologists.  Specifically, new approaches and
models are required for representing, storing, processing, and distributing
data produced by high-throughput methods.

\subsection{Scalability \& Performance}
\label{Scalability}

One of the key principles that has enabled the creation of high-throughput
methods is scalability.  Simply put, scalable methods are able to process large
amounts of data at the same level of performance as small amounts.  For
biological assays this is largely a problem of parallelization of chemical
reactions and miniaturization of devices.  In terms of computational
infrastructure, building effective, scalable systems also requires
parallelization.  This is often referred to by scalability engineers as
\emph{horizontal scaling} \cite{schlossnagle2006,arlitt2001}.

When setting up an infrastructure system using the \emph{horizontal scaling}
pattern, computer resources can be treated as groups of resources, or
\emph{clusters}.  Each cluster is responsible for one or more types of tasks,
and each component of the cluster is uniform.  This allows the throughput of
the cluster to be scaled simply by adding or removing components.

%See Also
genelogic		\cite{genelogic}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Informatics}
%@ The role of informatics in biology is dramatically changing the discipline
%and opening up the applications of high-throughput, multiplexing technology
The field of biology is in the process of rapidly changing from a discipline
that largely focuses on bottom up approaches to one that produces and analyzes
vast amounts of high-throughput, high-dimensional data. New emerging
technologies such as DNA microarrays, mass spectrometry, and rapid genome
sequencing mean that a biologist can now study entire organisms rather than just
one or two genes at a time.  This transition has created exciting opportunities
for integrating a wide variety of previously separate research areas including
engineering, statistics, computer science, and other disparate subjects.  The
end result is a field that is swiftly adopting new technologies and approaches
to analyze these highly integrated and challenging datasets. 

%@ Biologists today are facing a huge problem of data sharing, organizing,
%analysing, etc.  Needs solutions to a host of problems.
% Biologists today face huge problems of storing data and organizing large
% datasets such as microarrays.  The typical microarray hybridization experiment
% on the Affymetrix U133A platform contains approximately 22,000 features and
% individual raw image files can take up to 150Mb each of storage space each.
% Pre-processed data files that summarize image files take up less space but
% large datasets can still become difficult to work with.  The Celsius project,
% for example, mirrors public microarray respositories allowing for research into
% comparisons between microarrays from thousands of experiments.  The processed
% expression file for this conglomerate dataset is over 5Tb which is
% prohibatively large for most research labs to duplicate. Furthermore, the
% actual analysis of very large datasets from mass spec or microarray
% technologies can tax computer RAM and CPU capabilities.  The all-by-all
% comparison of probeset correlation, for example, is currently intractable on
% the Celsius dataset on current hardware without first dividing the task into
% smaller steps. For problems where a natural parellel solution is not possible
% limitations of CPU and RAM during analysis become acute issues.

\subsection{Informatic Challenges in Biology}

%@ These core problems break down into several related but distinct activites
% The informatics problems facing biologists can be divided into several distinct
% classes. Issues of working with microarray data can be thought of
% as a microacosum of the common types of problems biologist are experiencing as
% a result of new technologies.  A
% scientist studying gene expression patterns with a large number of microarrays
% experiments must be able to store both the raw and processed results.  These
% might include image files, processed expression data, normalized gene
% expression values, and other derived data.  Experimental annotations must be
% collected that clearly identify samples, how they were treated, and the
% parameters of the physical assay. Next, techniques for analyzing the data must
% be standardized.  This could include common algorithms or best practices for
% working with the data.  In microarrays, many competing analysis techniques are
% available and, regardless of the choice, a common way of accessing them is a key
% concern.  Furthermore, results can change with subsequent releases of new
% software or datasets. Since many scientific questions can take months or years
% to answer, keeping track of what resources were used to produce a given result
% are critical.  Microarray data can be normalized in a number of
% different ways with a variety of algorithms designed with similar but slightly
% different assumptions. Finally, sharing information is the final activity of
% most academic research.  Journals have become more sophisiticated in moving
% articles to completely digital distrubution and archiving models.  However as
% datasets continue to grow access via programatic interfaces is more and more
% important.  This allows for verification of existing results and also the
% synthesis of new questions and research directions.  Microarray data deposition
% in public repositories is a requirement for most journals yet automatic
% mechanisms for working with this data are lacking and requirements for what
% constitues primary data in this field are vague at best. The challenges facin

%@ Commonly applied solutions across the fields of information technology and
%computer science (collectively refered to as informatics) can meet many of
%these needs.  A hybrid approach is needed that blends the best resources from
%a variety of data driven desciplines with the highly parellel nature of modern
%genomics.
The issues facing biologists include storage, annotations, analysis, sharing, and
standardization.  All have roots in the highly technical and interdisciplinary
nature of modern biology.  Bioinformatics, computational biology, and other
related ``new'' fields take an interdisciplinary approach to solving these
challenges.  Commonly applied solutions across the fields of information
technology (IT), computer science, and engineering (collectively referred to
here as ``informatics'') can meet many of these needs.  A hybrid
approach that blends the best resources and practices from a variety
of data driven disciplines with the highly parallel nature of modern biology is needed.
Of particular note, the open source movement has achieved a high level of
success in creating free and flexible software that is philosophically aligned
with the principles of open, transparent, and freely accessible research.  It
is no wonder that most bioinformatics projects use open source software, such
as the Linux operating system, for the bulk of their work.  Solutions for many
of the technical challenges surrounding the transforming field of biology can
be found in the open source movement.

In the following sections, possible solutions to some of the specific informatics
challenges facing biology today are described. Summaries of the open source
bioinformatics projects in this dissertation are available in Table %XXX\ref{projects}.

%@ Informatic solutions borrowed from computer science include...
%@ These solutions can be adapted to the core problems...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsection{Further Work}

XXX XXX XXX

%Chapter \XXXchapter\ provides additional details about the application
%of logic analysis to microarray data.  This includes work supporting the claim
%that logic analysis is able to identify non-random gene expression
%relationships.  Chapter \XXXchapter\ explores the use of logic analysis
%as a general mechanism to discover meaningful gene relationships in additional
%microarray datasets.  It compares the performance of logic analysis to other
%feature selection algorithms and assess the ability of these features selected by
%logic analysis to classify samples in a supervised learning test with a variety
%of classification algorithms.


\section{Placeholder Section}
\label{XXX}

\bibliography{used}
%\bibliography {affy,scholar,pmid,thesis,la,dboutz}    % bibliography references
%\bibliographystyle {uclathes}
\bibliographystyle {plain}

